---
title: "Discussion5_Summarizing Dynamic Model Output"
format: revealjs
theme: solarized
resources: ["img/"]
css: ["slides.css"]
editor: visual
---

## Adding Carrying Capacity

## R code with carrying capacity {.scrollable}

```{r exampleode}
library(here)
library(deSolve)
library(tidyverse)

source(here("R/dexppop_play.R"))

dexppop_play

# create parameter list
initalrabbits = 2
years = seq(1,100,by=2)

newparms = list(r = .03, carry_capacity = 300)
# apply solver
results <- ode(initalrabbits, years, dexppop_play, newparms)
head(results)

# add more meaningful names
colnames(results) <- c('year','p')
```

## Plot

```{r, plotodep}
# plot
ggplot(as.data.frame(results),
       aes(x = year,
           y = p))+
  geom_point()+
  labs(y = 'Population',
       x = 'Year')
```

## Try again with different parameters {.scrollable}

```{r exampleode2}
# same initial condtions

initalrabbits = 2
years = seq(1,100,by=2)


# try again with different parameters
newparms = list(r = .04, carry_capacity = 500)

results2 <- ode(initalrabbits, years, dexppop_play, newparms)

colnames(results2) <- c('year','p_parm2')

# look at results
head(results2)

# plot

ggplot(as.data.frame(results2),
       aes(x = year,
           y = p))+
  geom_point()+
  labs(y = 'Population',
       x = 'Year')

# compare by combining into a single data frame
both <- inner_join(as.data.frame(results), as.data.frame(results2))

#both_p <- both %>% gather(key = "model", value = "P", -year)

both_p <- both %>% pivot_longer(-year, names_to = "model", values_to = "P")

```

## Plot

```{r, plotb}
# plot
ggplot(both_p, aes(x = year,y = P, col = model)) +
  geom_point() +
  labs(y = "Population", 
       x = "years")
```

## Extracting information from space-time results {.scrollable}

In class we've talking about summarizing results from dynamic models

-   pictures can be hard to interpret

-   summarizing over one of the dimensions (either space or time) can help

-   looking at a single trajectory through time

-   looking at spatial variation for one point in time

-   looking at spatial variation for multiple points in time

## Extracting from our diffusion model {.scrollable}

```{r}
source(here("R/diffusion.R"))
resfast_fixtime <- diff1(initialC = 100,
                         nx = 10,
                         dx = .5,
                         nt = 100,
                         dt = .5,
                         D = .2,
                         area = 1)

filled.contour(resfast_fixtime$conc, 
               xlab = 'time',
               ylab = 'distance')
```

# Summarizing Model Output

-   Summarize relative to what you care about

Lets say this was a pollutant injected into a stream...

-   lets say you have a vulnerable habit located 3 spatial units from the injection point

-   lets see what its pollutant exposure looks like.\

```{r one option}
# View(resfast_fixtime$conc)


# graph a single point in space through time
# single column (time)
plot(resfast_fixtime$conc[,8], ylab = 'Conc. after 3 spatial time units')
```

-   what if we were concerned about all locations

```{r all locations}
# plot all trajectories
# add a time column to concentration data frame and transform for plotting
resl <- as.data.frame(resfast_fixtime$conc) %>% 
  mutate(time = seq(1,100,by =1)) %>% 
  pivot_longer(-time, names_to = 'distance',
               values_to = 'conc')
ggplot(resl, aes(x = time, y = conc, col = distance))+
  geom_line()

# plot all places at each point in time
ggplot(resl, aes(time, conc, group = time)) +
  geom_boxplot()
```

# Spatial summaries

-   interested in how quickly the pollutant evens out through space

```{r spatial summaries}

# use apply to calculate the spatial variation for each row (e.g for each time point)
cvar <- resfast_fixtime$conc %>% 
  apply(1,var)

cmean <- resfast_fixtime$conc %>% 
  apply(1,mean)

spatial_aver = cbind.data.frame(cvar, cmean, time = seq(1,100))
# notice its the same as the number of time units (nt) used above

# plot spatial variation through time
ggplot(spatial_aver, aes(x = time, y = cvar))+
  geom_line() +
  labs(y = "Spatial Variation")

# plot coefficient of variation (so standard deviation divided by the mean)
ggplot(spatial_aver, aes(x = time, y = 100 * sqrt(cvar) / cmean))+
  geom_line() +
  labs(y = "Spatial Variation")
```

## it is helpful to turn the summary into a function that you can apply to the dynamic models output

## In class exercise {.scrollable}

For population model - write a function that summarizes something about the population over time
