---
title: "Assignment 6: Sobol Sensitivity Analysis with an ODE"
format: pdf
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(deSolve)
library(sensitivity)
library(tidyverse)
```

## Forest Growth Model

The forest growth ODE is defined as:

- If $C < \text{canopy\_thresh}$: $\frac{dC}{dt} = r \cdot C$ (exponential growth)
- If $C \geq \text{canopy\_thresh}$: $\frac{dC}{dt} = g \cdot \left(1 - \frac{C}{K}\right)$ (linear growth toward carrying capacity)

```{r}
source(here::here("forest_growth.R"))
```

## Part 1: Run Model for 300 Years

```{r}
# Parameters
params <- list(
  r             = 0.01,   # exponential growth rate (pre-canopy closure)
  g             = 2,      # linear growth rate (post-canopy closure), kgC/year
  K             = 250,    # carrying capacity, kgC
  canopy_thresh = 50      # canopy closure threshold, kgC
)

# Initial conditions and time
C0    <- c(C = 10)        # initial forest size, kgC
times <- seq(0, 300, by = 1)

# Solve ODE
result <- ode(y = C0, times = times, func = forest_growth, parms = params)
result_df <- as.data.frame(result)
```

## Part 2: Graph Trajectory

```{r}
ggplot(result_df, aes(x = time, y = C)) +
  geom_line(color = "forestgreen", linewidth = 1.2) +
  geom_vline(xintercept = result_df$time[which.min(abs(result_df$C - 50))],
             linetype = "dashed", color = "brown", alpha = 0.6) +
  annotate("text", x = result_df$time[which.min(abs(result_df$C - 50))] + 5,
           y = 20, label = "Canopy closure\nthreshold", color = "brown", size = 3) +
  labs(
    title    = "Forest Growth Over 300 Years",
    x        = "Time (years)",
    y        = "Forest Size (kgC)"
  ) +
  theme_bw()
```

## Part 3: Sobol Sensitivity Analysis

We explore how maximum forest size over 300 years varies with uncertainty in:

- `r`: pre-canopy exponential growth rate
- `g`: post-canopy linear growth rate
- `K`: carrying capacity
- `canopy_thresh`: canopy closure threshold

All parameters are assumed normally distributed with mean = baseline and SD = 10% of mean.

```{r}
# Number of samples
n_samples <- 1000

# Means and SDs (10% of mean)
means <- c(r = 0.01, g = 2, K = 250, canopy_thresh = 50)

sds   <- means * 0.10

# Generate two independent sample matrices
set.seed(42)
X1 <- as.data.frame(mapply(function(m, s) rnorm(n_samples, m, s),
                            means, sds))

X2 <- as.data.frame(mapply(function(m, s) rnorm(n_samples, m, s),
                            means, sds))

colnames(X1) <- colnames(X2) <- names(means)

# Create Sobol object
sobol_obj <- sobolSalt(model = NULL, X1 = X1, X2 = X2, scheme = "A", nboot = 100)
colnames(sobol_obj$X) <- names(params)
```

```{r wrapper}
p_wrapper <- function(r, 
                      g, 
                      K, 
                      canopy_thresh, 
                      initial_forest, 
                      simtimes, 
                      odefunc, 
                      metricfunc) {
  
  parms <- list(r = r, 
                g = g, 
                K = K, 
                canopy_thresh = canopy_thresh)
  
  result <- ode(y = initial_forest, times = simtimes, func = odefunc, parms = parms, method = "euler")
  
  result <- data.frame(time = result[, 1], C = result[, 2])
  
  return(metricfunc(result))
}


```

```{r}
# Set max forest function 
compute_max_forest <- function(result) {
  max(result$C)
}

# Run wrapper over all Sobol parameter sets
max_forest <-pmap(as.data.frame(sobol_obj$X), p_wrapper,
                initial_forest = c(C = 10),
                simtimes = seq(0, 300, by = 1),
                odefunc = forest_growth,
                metricfunc = compute_max_forest)


# Tell Sobol about the outputs
tell(sobol_obj, unlist(max_forest))
```

## Part 4: Sensitivity Analysis Results

```{r}
# Extract indices
S  <- sobol_obj$S   # first-order
T  <- sobol_obj$T   # total-order

cat("First-order Sobol indices (S):\n")
print(S)
cat("\nTotal-order Sobol indices (T):\n")
print(T)
```

```{r fig.cap="Distribution of maximum forest size (kgC) across Sobol parameter samples."}
plot_df <- data.frame(max_C = as.numeric(max_forest))

# Boxplot of max forest size
plot_df %>% 
  ggplot(aes(x = "", y = max_C)) +
  geom_boxplot(fill = "lightgreen", color = "darkgreen", outlier.alpha = 0.3) +
  labs(
    title = "Sobol Sensitivity: Maximum Forest Size over 300 Years",
    x     = NULL,
    y     = "Maximum Forest Size (kgC)"
  ) +
  theme_bw()
```

```{r sobol-barplot, fig.cap="First-order and total-order Sobol sensitivity indices for each parameter."}
# Bar plot of indices
indices_df <- bind_rows(
  S %>%  mutate(index = "First-order (S)",  param = rownames(S)),
  T %>%  mutate(index = "Total-order (T)", param = rownames(T))
)

ggplot(indices_df, aes(x = param, y = original, fill = index)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = `min. c.i.`, ymax = `max. c.i.`),
                position = position_dodge(0.9), width = 0.25) +
  scale_fill_manual(values = c("steelblue", "coral")) +
  labs(
    title = "Sobol Sensitivity Indices",
    x     = "Parameter",
    y     = "Sobol Index",
    fill  = "Index Type"
  ) +
  theme_bw()
```

## Part 5: Extra Credit — Forest Size at 100 Years

```{r}
compute_forest_100 <- function(result) {
  result$C[result$time == 100]
}

sobol_100 <- sobolSalt(model = NULL, X1 = X1, X2 = X2, scheme = "A", nboot = 100)
colnames(sobol_100$X) <- c("r", "g", "K", "canopy_thresh")

results_100 <- pmap(as.data.frame(sobol_100$X), p_wrapper,
                    initial_forest = c(C = 10),
                    simtimes = seq(0, 100, by = 1),
                    odefunc = forest_growth,
                    metricfunc = compute_forest_100)

colnames(sobol_100$X) <- c("r", "g", "K", "canopy_thresh")
tell(sobol_100, unlist(results_100))

print(sobol_100$S)
print(sobol_100$T)
```

## Discussion

The sensitivity analysis reveals that **carrying capacity (K)** and the **post-canopy growth rate (g)** have the largest influence on maximum forest size, which makes intuitive sense since K is the ceiling and g controls how quickly the forest approaches it once canopy closure is reached. In contrast, the pre-canopy exponential rate `r` has relatively low sensitivity because forests spend less time in that phase over a 300-year horizon.

From a climate perspective, this matters because rising temperatures and shifting precipitation could alter both `g` (net productivity of closed-canopy forests) and `K` (the carbon carrying capacity, which depends on soil quality and water availability). If climate stress reduces `K` or `g`, our models would systematically overestimate long-term carbon sequestration potential, with important implications for carbon accounting and forest-based climate mitigation strategies.
```
