---
title: "Assignment 7"
format: html
editor: visual
author: Peter Vitale
date: '2026-03-09'
cache: TRUE
---

## Import source function and packages 

```{r}
pacman::p_load(
  'deSolve',
  'here',
  'tidyverse',
  'sensitivity'
)

source(here::here('R', 'lotvmodK.R'))
```


## 1. Run the model

```{r}
#| warning: false
#| message: false
#| results: hide


times <- seq(0, 50, by=1)

hunt_levels <- seq(0, 200, by=10)

# Parameters
pars <- c(rprey=0.95, 
          alpha=0.01, 
          eff=0.6, 
          pmort=0.4, 
          K=2000, 
          min_prey=100)

currpop <- c(prey=500, pred=50)

# Solve ODE
results <- lapply(hunt_levels, function(h) { # Set hunting levels
  pars["hunt"] <- h # append different levels to hunt
  # Run ODE
  out <- ode(y=currpop, times=times, func=lotvmodK, parms=pars)
  #Make sure its a dataframe 
  as.data.frame(out)
})
```

## 2. Compute Stability 
```{r}
compute_stability <- function(result) {
  after10 <- result[result$time > 10, "prey"]
  if (length(after10) == 0) return(0)
  min(after10)
}

stability <- sapply(results, compute_stability)

stability_df <- data.frame(hunt = hunt_levels, min_prey = stability)
```

## 3. Plot outcomes 
```{r}
ggplot(stability_df, aes(x=hunt, y=min_prey)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept=100, color="firebrick", linetype="dashed") +
  labs(x="Hunting Level", y="Minimum Prey Population (after year 10)",
       title="Effect of Hunting on Prey Stability")+
  theme_minimal()
```

## 4. Adding uncertainty

Now that we have the basic framework, we can add uncertainty. I chose this hunting metric as hunting is usually regulated to a specific amount, so having hunting be a quantity instead of a proportion made the most sense to me.  

```{r}
#| warning: false
#| message: false
#| results: hide


n_sims <- 100

uncertainty_results <- lapply(1:n_sims, function(i) {
  # Vary parameters slightly around best guesses
  pars_i <- c(
    rprey = rnorm(1, mean=0.95, sd=0.1),
    alpha = rnorm(1, mean=0.01, sd=0.001),
    eff   = rnorm(1, mean=0.6,  sd=0.05),
    pmort = rnorm(1, mean=0.4,  sd=0.05),
    K     = 2000,
    min_prey = 100
  )
  
  # Steps 1-2 for this parameter set
  sapply(hunt_levels, function(h) {
    pars_i["hunt"] <- h
    
  
  out <- ode(y=currpop, times=times, func=lotvmodK, parms=pars_i)
  
    compute_stability(as.data.frame(out))
  })
})
```

## 5. Plotting Uncertainty 

```{r}
uncertainty_df <- do.call(rbind, lapply(1:n_sims, function(i) {
  data.frame(hunt=hunt_levels, min_prey=uncertainty_results[[i]], sim=i)
}))

ggplot(uncertainty_df, aes(x=hunt, y=min_prey, group=sim)) +
  geom_line(alpha=0.1) +
  geom_hline(yintercept=100, color="firebrick", linetype="dashed") +
  labs(x="Hunting Level", y="Minimum Prey Population",
       title="Stability Under Parameter Uncertainty")+
  theme_minimal()
```
## 6. Sobol Sensitivity Analysis
```{r}
#| warning: false
#| message: false
#| results: hide
n <- 1000
X1 <- data.frame(
  rprey = rnorm(n, mean=0.95, sd=0.1),
  alpha = rnorm(n, mean=0.01, sd=0.001),
  eff   = rnorm(n, mean=0.6,  sd=0.05),
  pmort = rnorm(n, mean=0.4,  sd=0.05)
)
X2 <- data.frame(
  rprey = rnorm(n, mean=0.95, sd=0.1),
  alpha = rnorm(n, mean=0.01, sd=0.001),
  eff   = rnorm(n, mean=0.6,  sd=0.05),
  pmort = rnorm(n, mean=0.4,  sd=0.05)
)
sens <- sobolSalt(model=NULL, X1, X2, scheme="A", nboot=100)

run_model <- function(params) {
  apply(params, 1, function(p) {
    pars_i <- c(
      rprey = max(0.01, p[1]),
      alpha = max(0.001, p[2]),
      eff   = max(0.01, p[3]),
      pmort = max(0.01, p[4]),
      K=2000, min_prey=100, hunt=50
    )
    out <- tryCatch(
      ode(y=currpop, times=times, func=lotvmodK, parms=pars_i),
      error = function(e) NULL
    )
    if (is.null(out)) return(0)
    compute_stability(as.data.frame(out))
  })
}

y <- run_model(sens$X)
tell(sens, y)
```

## 7. Plot Sobol Results
```{r}
#| warning: false
#| message: false
sobol_df <- data.frame(
  parameter = c("rprey", "alpha", "eff", "pmort"),
  S1 = sens$S[, "original"],
  T1 = sens$T[, "original"]
) %>%
  pivot_longer(cols = c(S1, T1), names_to = "index", values_to = "value")

ggplot(sobol_df, aes(x=parameter, y=value, fill=index)) +
  geom_col(position="dodge") +
  labs(x="Parameter", y="Sensitivity Index",
       fill="Index",
       title="Sobol Sensitivity Indices") +
  theme_minimal()
```

## 8. Discussion 

alpha and pmort most strongly influence prey stability, suggesting that predator-prey interaction rate and predator mortality are the key drivers of system behavior under hunting pressure. rprey has negligible direct influence but contributes through interactions.

